Compositional Certified Resource Bounds\ifshort{\vspace{-1.15cm}}{}

### 2014-11-13 paper/main.tex
minor
typos

This paper presents a new compositional framework for automatically
deriving resource bounds on C programs.  This new
approach is an attempt to unify the two aforementioned
categories: It solves the compositionality issues of techniques
for numerical imperative code by adapting amortized-analysis--based
techniques from the functional world.
%
Our automated analysis is able to infer
resource bounds on C programs with mutually-recursive functions and integer loops.
%
The resource behavior of functions can be summarized
in a function specification that can be used at every
call site without accessing the function body.
%
To our knowledge this is the first technique based on
amortized analysis that is able to derive bounds that depend on negative numbers
and differences of variables.  It is also the first
resource analysis technique for C that deals naturally with
recursive functions and sequenced loops, and can handle resources that can
become available during execution (e.g., when freeing
memory).  Compared to more classical
approaches based on ranking functions, our tool inherits
the benefits of amortized reasoning.  Using only one
simple mechanism, it handles:
\itemskip
\begin{itemize}
\item interactions between \emph{sequential loops or
  function calls} through size changes of variables,
\itemskipIn
\item \emph{nested loops} that influence each other
  with the same set of modified variables,
\itemskipIn
\item and \emph{amortized bounds} as found, for example, in
  the Knuth-Morris-Pratt algorithm for string search.
\end{itemize}
\itemskip
The main innovation that makes amortized analysis work
on imperative languages is to base the analysis on a
Hoare-like logic and to track multivariate quantities instead
of program variables.  This leads to precise bounds
expressed as functions of sizes $|[x, y]| = \max(0, y-x)$ of
intervals.
%
A distinctive feature of our analysis system is that
it reduces linear bound inference  to a linear optimization problem
that can be solved by off-the-shelf LP solvers.  This enables the
efficient inference of global bounds for larger programs.
%
Moreover, our local inference rules automatically generate
proof certificates that can be easily checked in linear time.

@@

We have implemented the analysis system in the tool
\toolname and experimentally evaluated its effectiveness by
analyzing system code and examples from the literature.
\toolname has automatically derived global resource bounds for
more than $\totalLoc$ lines of C code from the cBench benchmark suite.
%
\iffull{\pref{app:cat}}{The extended version
  of this article~\cite{anon_extended}}{} contains more than $30$
challenging loop and recursion patterns that we collected from open
source software and the literature.  Our analysis can find
asymptotically tight bounds for all but 1 of these patterns, and in
most cases the derived constant factors are tight.  To compare $\toolname$
with existing techniques, we tested our examples
with tools such as KoAT~\cite{BrockschmidtEFFG14},
Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
show that the bounds that we derive are often more precise than those
derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
uses amortization techniques, is able to achieve a similar precision.

@@

In contrast to the quantitative annotations, logical contexts $\Gamma$
in the presented inference system are left abstract.  This allows for
simpler rules and leaves room for future improvements.
Our implementation finds these logical contexts using abstract
interpretation with the domain of linear inequalities.  We observed that
the rules of the analysis often requires only minimal local knowledge.
This means that it is not necessary for us to compute precise loop
invariants and only a rough fixpoint (e.g. keeping only inequalities on
variables unchanged by the loop) is sufficient to obtain good bounds.

@@

\paraskip
\paragraph{Mutual Recursive Functions.}

As mentioned, the analysis also handles advanced control flow like
\code{break} and \code{return} statements, and mutual recursion.
Example \emph{t39} in \pref{fig:ex_list_2} contains two mutually-recursive functions with their
automatically derived tick bounds.  The function \code{c\_down}
decrements its first argument $x$ until it reaches the second argument
$y$. It then recursively calls the function \code{c\_up},
which is dual to \code{c\_down}.  Here, we
c up $y$ by $2$ and recursively call \code{c\_down}.  $\toolname$ is
the only available system that computes a tight
bound.  \iffull{The analysis amounts to computing the meeting
point of two trains that approach each other with different speeds.}{}
% This is comparable to the
% determination of the point at which to trains meet if they approach
% each other with different speeds.

@@

Example \emph{t61} in \pref{fig:ex_list_2} is typical for implementations
of block-based cryptographic primitives: Data of arbitrary length, is
consumed in blocks and the leftover is stored in a buffer for future use
when more data is available.  It is present in all the block encryption
routines of PGP and also used in performance critical
code to unroll a loop. For example we found it in a bit manipulating
function of libtiff library and a CRC computation routine of MAD,
an MPEG decoder.  This looping pattern is handled particularly well
by our method. If $N \geq 8$, $\toolname$ infers the bound
$\frac{N}{8} |[0,l]|$, but if $N<8$, it infers $7\frac{8-N}{8} +
\frac{N}{8}|[0,l]|$. The selection of the block size ($8$) and the cost
in the second loop (\code{tick(1)}) are random choices and $\toolname$
would also derive tight bound for other values.

@@

Example \emph{t62} (\pref{fig:ex_list_2})
is the inner loop of a quick sort implementation in cBench.  More
precisely, it is the partitioning part of the algorithm.  This partition
loop has linear complexity, and feeding into our analysis gives the
correct and tight worst-case bound $2+3|[l,h]|.$  With the amortized
analysis approach in mind, it is clear that the three loops get their
potential from the same interval $[l,h]$.  However, KoAT fails to
find a bound and LOOPUS derives the quadratic bound
$(h-l-1)^2$.  Following the classical technique, these tools try to find one
linear ranking-function for each loop and combine them multiplicatively
or additively.

@@

\iffull{}{ The main parts of the soundness proved are formalized with
  Coq and available for download. The full definitions of the cost
  semantics and the quantitative Hoare logic, and more details on the
  soundness proof can be found in the extended version of this
  article.}

>>>>

This paper presents a new compositional framework for automatically
deriving resource bounds on C programs.  This new
approach is an attempt to unify the two aforementioned
categories: It solves the compositionality issues of techniques
for numerical imperative code by adapting amortized-analysis--based
techniques from the functional world.
%
Our automated analysis is able to infer
resource bounds on C programs with mutually-recursive functions and integer loops.
%
The resource behavior of functions can be summarized
in a function specification that can be used at every
call site without accessing the function body.
%
To our knowledge this is the first technique based on
amortized analysis that is able to derive bounds that depend on negative numbers
and differences of variables.  It is also the first
resource analysis technique for C that deals naturally with
recursive functions and sequenced loops, and can handle resources that can
become available during execution (e.g., when freeing
memory).  Compared to more classical
approaches based on ranking functions, our tool inherits
the benefits of amortized reasoning.  Using only one
simple mechanism, it handles:
\itemskip
\begin{itemize}
\item interactions between \emph{sequential loops or
  function calls} through size changes of variables,
\itemskipIn
\item \emph{nested loops} that influence each other
  with the same set of modified variables,
\itemskipIn
\item and \emph{amortized bounds} as found, for example, in
  the Knuth-Morris-Pratt algorithm for string search.
\end{itemize}
\itemskip
The main innovations that make amortized analysis work
on imperative languages are to base the analysis on a
Hoare-like logic and to track multivariate quantities instead
of program variables.  This leads to precise bounds
expressed as functions of sizes $|[x, y]| = \max(0, y-x)$ of
intervals.
%
A distinctive feature of our analysis system is that
it reduces linear bound inference  to a linear optimization problem
that can be solved by off-the-shelf LP solvers.  This enables the
efficient inference of global bounds for larger programs.
%
Moreover, our local inference rules automatically generate
proof certificates that can be easily checked in linear time.

@@

We have implemented the analysis system in the tool
\toolname and experimentally evaluated its effectiveness by
analyzing system code and examples from the literature.
\toolname has automatically derived global resource bounds for
more than $\totalLoc$ lines of C code from the cBench benchmark suite.
%
\iffull{\pref{app:cat}}{The extended version
  of this article~\cite{anon_extended}}{} contains more than $30$
challenging loop and recursion patterns that we collected from open
source software and the literature.  Our analysis can find
asymptotically tight bounds for all but one of these patterns, and in
most cases the derived constant factors are tight.  To compare $\toolname$
with existing techniques, we tested our examples
with tools such as KoAT~\cite{BrockschmidtEFFG14},
Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
show that the bounds that we derive are often more precise than those
derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
uses amortization techniques, is able to achieve a similar precision.

@@

The abstract program state is represented as \emph{logical contexts} in the
derivation system used by our automated tool.
Our implementation finds these logical contexts using abstract
interpretation with the domain of linear inequalities.  We observed that
the rules of the analysis often requires only minimal local knowledge.
This means that it is not necessary for us to compute precise loop
invariants and only a rough fixpoint (e.g. keeping only inequalities on
variables unchanged by the loop) is sufficient to obtain good bounds.

@@

\paraskip
\paragraph{Mutually Recursive Functions.}

As mentioned, the analysis also handles advanced control flow like
\code{break} and \code{return} statements, and mutual recursion.
Example \emph{t39} in \pref{fig:ex_list_2} contains two mutually-recursive functions with their
automatically derived tick bounds.  The function \code{c\_down}
decrements its first argument $x$ until it reaches the second argument
$y$. It then recursively calls the function \code{c\_up},
which is dual to \code{c\_down}.  Here, we
count up $y$ by $2$ and call \code{c\_down}.  $\toolname$ is
the only available system that computes a tight
bound.  \iffull{The analysis amounts to computing the meeting
point of two trains that approach each other with different speeds.}{}
% This is comparable to the
% determination of the point at which to trains meet if they approach
% each other with different speeds.

@@

Example \emph{t61} in \pref{fig:ex_list_2} is typical for implementations
of block-based cryptographic primitives: Data of arbitrary length is
consumed in blocks and the leftover is stored in a buffer for future use
when more data is available.  It is present in all the block encryption
routines of PGP and also used in performance critical
code to unroll a loop. For example we found it in a bit manipulating
function of the libtiff library and a CRC computation routine of MAD,
an MPEG decoder.  This looping pattern is handled particularly well
by our method. If $N \geq 8$, $\toolname$ infers the bound
$\frac{N}{8} |[0,l]|$, but if $N<8$, it infers $7\frac{8-N}{8} +
\frac{N}{8}|[0,l]|$. The selection of the block size ($8$) and the cost
in the second loop (\code{tick(1)}) are random choices and $\toolname$
would also derive tight bound for other values.

@@

Example \emph{t62} (\pref{fig:ex_list_2})
is the inner loop of a quick sort implementation in cBench.  More
precisely, it is the partitioning part of the algorithm.  This partition
loop has linear complexity, and feeding into our analysis gives the
correct and tight worst-case bound $2+3|[l,h]|.$  With the potential
method in mind, it is clear that the three loops get their
potential from the same interval $[l,h]$.  However, KoAT fails to
find a bound and LOOPUS derives the quadratic bound
$(h-l-1)^2$.  Following the classical technique, these tools try to find one
linear ranking-function for each loop and combine them multiplicatively
or additively.

@@

\iffull{}{ The main parts of the soundness proof are formalized with
  Coq and available for download. The full definitions of the cost
  semantics and the quantitative Hoare logic, and more details on the
  soundness proof can be found in the extended version of this
  article.}
